---
output:
  knitrBootstrap::bootstrap_document:
    theme: readable
    highlight: zenburn
    theme.chooser: TRUE
    highlight.chooser: TRUE
  html_document:
    toc: true
    highlight: zenburn
---
```{r knitr,eval=F,echo=F}
library(rmarkdown)
library(knitrBootstrap)
knitr::opts_chunk$set(tidy=TRUE, highlight=TRUE, dev="png", fig.width=9,fig.heigh=9,
               cache=FALSE, highlight=TRUE, autodep=TRUE, warning=FALSE, error=FALSE,
               message=FALSE, prompt=TRUE, comment='', fig.cap='', bootstrap.show.code=FALSE)
render("~/repos/pcbc_c4_analysis/code/Rmd/DE_MatrixmiRNA_seqbuster.Rmd", output_dir = "~/Dropbox/Public/hsph/pcbc")
```

# metadata and covariates

```{r meta}
library(tidyr)
library(dplyr)
library(knitr)
library(synapseClient)
library(DESeq2)
library(edgeR)
library(vcd)
library(limma)
library(reshape2)
library(gplots)
library(ggplot2)
library(RColorBrewer)
library(corrplot)
library(psych)
library(pheatmap)

source("~/repos/pcbc_c4_analysis/code/R/lib/qc.R")
source("~/repos/pcbc_c4_analysis/code/R/lib/covariateAnalysis.R")

synapseLogin()

# Get metadata
metadata_id = "syn3219876"
metadata_obj = synTableQuery(paste('SELECT * FROM',metadata_id,sep=' '))
metadata = metadata_obj@values %>% distinct(UID)
row.names(metadata) = metadata$UID

mat_obj =  synGet(id="syn3355993")
mat = mat_obj@filePath
# mat = read.csv("~/repos/pcbc_c4_analysis/data/mirna.csv", check.names = FALSE)

reads_obj =  synGet(id="syn3355994")
reads = reads_obj@filePath
# reads = read.table("~/repos/pcbc_c4_analysis/data/smallRNA_reads.tsv", check.names = FALSE, row.names=1, header=T)

keep = ( (reads %>% mutate(samples=rownames(reads)) %>% filter(mirna > 500000 & norm > 0.2))[,"samples"] )
metadata$size = reads[as.character(row.names(metadata)),"norm"]
metadata$size_cat = reads[as.character(row.names(metadata)),"breaks"]
metadata$size_cat = reads[as.character(row.names(metadata)),"breaks"]
# join meso-15 and meso-30
stages =  gsub("-5","Early",gsub("-30","Late",gsub("-15","Late",metadata$Diffname_short)))
metadata$Diffname_short = as.character(stages)
```

## Clean and sync tables

Just make sure the tables are sync, same order for samples

```{r clean_metadata}
meta_clean = metadata[colnames(mat), ]
meta_clean[meta_clean == 'N/A'] = NA
meta_clean[is.na(meta_clean)] = 'unk'
meta_clean[meta_clean == ''] = 'unk'
```

```{r clean}
meta_clean = meta_clean[keep,]
mat = mat[,keep]

keep_meta = meta_clean$Diffname_short != "unk"
meta_clean = meta_clean[keep_meta,]
mat = mat[,keep_meta]

meta_clean$mirna= log10(colSums(mat))

new_factors = c("Diffname_short", "Tissue_of_Origin","run","Originating_Lab_ID")
```

# Filtering by group

I want to try if filtering by stable miRNAs give better signal. 
I keep miRNAs that are express >= 5 (after normalization) in 90% of the samples of each group.
That would mean to filtering by group and the put all miRNAs together.

```{r sc}
myPalette <- colorRampPalette(rev(brewer.pal(11, "Spectral")))
sc <- scale_fill_gradientn(colours = myPalette(100), limits=c(1, 20))

clean_table = function(what, mat, meta_clean, new_factors){
    i = meta_clean %>% filter(Diffname_short %in% what)
    m = mat[,i$UID]
    c = do.call(paste0,i[,new_factors])

    d = data.frame(row.names = colnames(m), cov=c)
    dds = DESeqDataSetFromMatrix(m,colData = d,design = ~ 1)
    dds = estimateSizeFactors(dds)
    mat_norm = counts(dds, normalized=T)
    cutoff = round(ncol(mat_norm)*0.9)
    clean = as.data.frame((mat_norm[rowSums(mat_norm>5)>cutoff,]))
    list(mat=clean, meta=i)
}

plot_density = function(clean, i){
    
    #voom_mat = voom(mat_norm, design = design, plot=T,lib.size=size_factor)
    #corrected = voom_mat$E / voom_mat$weight
    
    x = melt(log2(clean+1) %>% mutate(mir=rownames(clean)))
    idx = match(x$variable, i$UID)
    x$type = i[idx,"Diffname_short"]
    x$x_jitter = 1
    print( ggplot(x %>% filter(value>0), aes(x=value,group=variable,colour=type)) + geom_line(aes(color=type), stat="density",  alpha=0.4) + xlab("expression")+facet_wrap(~variable) )
    
    idx = sort(rowMeans(clean),index.return =TRUE)
    x$mir = factor(x$mir, levels=row.names(clean)[idx$ix])
    print(ggplot(x) + 
        geom_tile(aes(x=variable, y=mir, fill=value)) +
        sc +
        scale_x_discrete(labels="") +
        scale_y_discrete(labels="") )
}

sc_clean = clean_table("SC",mat, meta_clean, new_factors)
mesoearly_clean = clean_table("MESOEarly",mat, meta_clean, new_factors)
mesolate_clean = clean_table("MESOLate",mat, meta_clean, new_factors)
ecto_clean = clean_table("ECTO",mat, meta_clean, new_factors)
eb_clean = clean_table("EB",mat, meta_clean, new_factors)
de_clean = clean_table("DE",mat, meta_clean, new_factors)

mir_sub = unique(c(row.names(sc_clean$mat),
                   row.names(ecto_clean$mat),
                   row.names(eb_clean$mat),
                   row.names(de_clean$mat),
                   row.names(mesolate_clean$mat),
                   row.names(mesoearly_clean$mat)))
samples_sub = unique(c(colnames(sc_clean$mat),
                       colnames(ecto_clean$mat),
                       colnames(eb_clean$mat),
                       colnames(de_clean$mat),
                       colnames(mesolate_clean$mat),
                       colnames(mesoearly_clean$mat)))


mat_sub = mat[mir_sub, samples_sub]
meta_sub = meta_clean[samples_sub,]
# covariates = do.call(paste0,meta_sub[,new_factors])
# qc_figures(mat_sub, meta_sub, covariates)
```

### heatmap with voom normalization

```{r voom-norm}
voom_mat = voom(mat_sub, design = model.matrix(~ 0 + ., model.frame(~ 0 + . , meta_sub[,new_factors])), plot=T)
pheatmap(voom_mat$E, clustering_distance_rows = "correlation", clustering_distance_cols = "correlation", clustering_method = "ward", annotation = meta_sub[,c("Diffname_short","Cell_Line_Type","size","mirna_b"),drop=F], show_rownames = F)
```

### heatmap with rlog-deseq normalization

```{r deseq-norm}
d = meta_sub[,new_factors]
dds = DESeqDataSetFromMatrix(mat_sub,colData = d,design = ~ Diffname_short)
rlog_mat = rlog(dds)
rlog_Mat = assay(rlog_mat)
rlog_Mat[rlog_Mat > 20]=20
meta_sub$mirna_b = cut(meta_sub$mirna, breaks=c(0,6,6.6,9))
pheatmap(rlog_Mat, clustering_distance_rows = "correlation", clustering_distance_cols = "correlation", clustering_method = "ward", annotation = meta_sub[,c("Diffname_short","Cell_Line_Type","size","mirna_b"),drop=F], show_rownames = F)
```

## Non negative matrix factorization

Using this package to find modules( sets of mirnas that explain the different groups.)

http://cran.r-project.org/web/packages/NMF/vignettes/NMF-vignette.pdf

```{r nmf}
library(NMF)
res_nmf = nmf(log2(rlog_Mat^2+2), 3:7)
consensusmap(res_nmf, annCol = meta_sub[,c("Diffname_short","Cell_Line_Type")])
plot(res_nmf)
res_nmf_4 = nmf(log2(rlog_Mat^2+2), 4)
coefmap(res_nmf_4,annCol = meta_sub[,c("Diffname_short","Cell_Line_Type")])
basismap(res_nmf_4, subsetRow=TRUE)
consensusmap(res_nmf_4, annCol = meta_sub[,c("Diffname_short","Cell_Line_Type")])

# metagene
s = featureScore(res_nmf_4)
s = extractFeatures(res_nmf_4)
nmf_select = rownames(rlog_Mat[unique(unlist(s)),])
# X <- syntheticNMF(100, 3, 20, noise = 2)

pheatmap(rlog_Mat[nmf_select,], clustering_distance_rows = "correlation", clustering_distance_cols = "correlation", clustering_method = "ward", annotation = meta_sub[,c("Diffname_short","Cell_Line_Type","size","mirna_b"),drop=F], show_rownames = F)
```

It worked without the MESOLate samples, no quite well detecting modules with all groups.

number of differents miRNAs in modules `r length(unlist(s))`

## Differential expression

```{r limma}

meta_sub$Originating_Lab_ID = as.numeric(as.factor(meta_sub$Originating_Lab_ID))
meta_sub$run = as.numeric(as.factor(meta_sub$run))
meta_sub$Tissue_of_Origin = gsub(" ",".",meta_sub$Tissue_of_Origin)

dge <- DGEList(counts=mat_sub)
dge <- calcNormFactors(dge)
dm = model.matrix(~ 0 + ., model.frame(~ 0 + . , meta_sub[,new_factors]))
v <- voom(dge,dm, plot=TRUE)
fit <- lmFit(v,dm)
my.contrast = makeContrasts( ECTO.DE=Diffname_short.ECTO - Diffname_short.DE, 
                             ECTO.SC=Diffname_short.ECTO-Diffname_short.SC,
                             ECTO.EB=Diffname_short.ECTO-Diffname_short.EB,
                             ECTO.MESOEarly=Diffname_short.ECTO-Diffname_short.MESOEarly,
                             SC.DE=Diffname_short.SC - Diffname_short.DE, 
                             SC.EB=Diffname_short.SC-Diffname_short.EB,
                             SC.MESOEarly=Diffname_short.SC-Diffname_short.MESOEarly,
                             EB.DE=Diffname_short.EB - Diffname_short.DE, 
                             EB.MESOEarly=Diffname_short.EB-Diffname_short.MESOEarly,
                             MESOEarly.DE = Diffname_short.MESOEarly - Diffname_short.DE,
                             MESOEarly.MESOLate = Diffname_short.MESOEarly - Diffname_short.MESOLate,
                             MESOLate.DE = Diffname_short.MESOLate - Diffname_short.DE,
                             levels=dm$design)
fit = contrasts.fit(fit, my.contrast)
fit <- eBayes(fit)

top = list()
for (c in colnames(my.contrast)){
    top[[c]] = topTable(fit, coef=c,number=Inf) %>% mutate(mirs=rownames(fit$coefficients)) %>% filter(adj.P.Val<0.01, abs(logFC)>2)   %>% mutate(comparison=c)  
}

data.frame(sapply(top, nrow))

res = do.call(rbind, top)
# write.table(res, "~/repos/pcbc_c4_analysis/data/DE_mirnas.tsv", row.names=F, quote=F, sep="\t") 
```


unique miRNAs `r length(unique(unlist(sapply(top, function(x){x$mirs}))))`

```{r limma-heatmap}
select = as.character(unique(unlist(sapply(top, function(x){x$mirs}))))

pheatmap(rlog_Mat[select,], clustering_distance_rows = "correlation", clustering_distance_cols = "correlation", clustering_method = "ward", annotation = meta_sub[,c("Diffname_short","Cell_Line_Type","size","mirna_b"),drop=F], show_rownames = F)
```


Common miRNAs between voom-limma and NMF are noisy signal

```{r common}
common = intersect(nmf_select, select)
pheatmap(rlog_Mat[common,], clustering_distance_rows = "correlation", clustering_distance_cols = "correlation", clustering_method = "ward", annotation = meta_sub[,c("Diffname_short","Cell_Line_Type","size","mirna_b"),drop=F], show_rownames = F)
```


