---
title: "miRNA-mRNA network analysis with mixed effect modilng"
author: "Lorena Pantano"
date: "`r date()`"
output: html_document
---

```{r knitr,eval=F,echo=F}
library(rmarkdown)
library(knitrBootstrap)
#render("~/repos/pcbc_c4_analysis/code/Rmd/miRNA_mRNA_network.Rmd", output_dir = "~/Dropbox/Public/hsph/pcbc")
```


```{r load}

library(tidyr)
library(dplyr)
library(knitr)
library(synapseClient)
library(limma)
library(reshape2)
library(gplots)
library(ggplot2)
library(RColorBrewer)
library(knitr)
library(NMF)

synapseLogin()

knitr::opts_chunk$set(tidy=TRUE, highlight=TRUE, dev="png", fig.width=9,fig.heigh=9,
               cache=FALSE, highlight=TRUE, autodep=TRUE, warning=FALSE, error=FALSE,
               message=FALSE, prompt=TRUE, comment='', fig.cap='', bootstrap.show.code=FALSE, echo=FALSE)

mirna_pval_obj =  synGet(id="syn4596141")
mirna_pval = read.table(mirna_pval_obj@filePath, check.names = F, header=T, row.names=1, sep="\t")
mirna_logfc_obj =  synGet(id="syn4595547")
mirna_logfc = read.table(mirna_logfc_obj@filePath, check.names = F, header=T, row.names=1, sep="\t")
```

```{r fn}
.format_table = function(m_logfc, m_pval, FCl=0.5, PVl=0.05){
    # m_logfc = .grep(m_logfc, comp)
    # m_pval = .grep(m_pval, comp)
    
    m_sig = lapply(names(m_pval), function(x){
      sig = m_pval[,x] < PVl & abs(m_logfc[,x]) > FCl
      if (sum(sig)>0){
        dt = data.frame(fc=m_logfc[sig, x],gene=rownames(m_logfc)[sig])
      
        dt$cont = x
        dt
      }
    })
    Filter(is.null, m_sig)
    do.call(rbind, m_sig)
}

filter_deg = function(th, mirna_pval, mirna_logfc ){
  FCl=th[1]
  PVl=th[2]
  dt = .format_table(mirna_logfc, mirna_pval, FCl, PVl)
  dt$pval=PVl
  dt$logFC=FCl
  dt
}

```

Number of differential expressed genes with different cutoffs. Number of genes are normalized to the comparison that showed the maximum number of de-regulated genes. See table below.

```{r loop, results="asis"}
th = expand.grid(c(0,0.5,1,2),c(0.01,0.05,0.1))
t = apply(th, 1,function(x) {filter_deg(x,mirna_pval, mirna_logfc)})

comp = do.call(rbind, t)
comp = left_join( comp %>% group_by(cont, pval, logFC) %>% dplyr::summarise(total=n()) %>%
                       mutate(cutoffs=paste0("logFC:",logFC," FDR:", pval), 
                              comparison=substr(cont, 1, 25)),
                  comp %>% group_by(cont, pval, logFC) %>% dplyr::summarise(maxn=n()) %>% 
                    group_by(cont) %>% dplyr::summarise(maxn=max(maxn)), by='cont' )  %>%
       mutate(prop=total/maxn)

kable(comp[!duplicated(comp$cont),c("cont","maxn")])

```


```{r fig-loop, fig.width=11}

ggplot(comp, aes(y=cutoffs, x=comparison, fill=prop)) +
  geom_tile() +
  scale_fill_gradient2(low = "white", midpoint = 0.5, mid = "grey", high = "black") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```

How the different cutoffs affect clustering. Using just the voom normalization matrix with the design that was used to do the differential expression. Distance calculated with correlation and clustering with ward method.

```{r heatmaps, fig.height=11, fig.width=11}

metadata_id = "syn3219876"
metadata_obj = synTableQuery(paste('SELECT * FROM',metadata_id,sep=' '))
metadata = metadata_obj@values %>% distinct(UID)
row.names(metadata) = metadata$UID

mirna_rlog_obj =  synGet(id="syn4595556")
rlog_Mat = read.table(mirna_rlog_obj@filePath, check.names = F, header=T, row.names=1, sep="\t")


.grep <- function(df, comp){
  idx = rowSums( sapply(paste0('_',comp,'_'), function(x)grepl(x, df$cont) ) ) > 0
  df[idx,]
}

seen = c("EB", "DE", "SC", "MESO5", "MESO15", "MESO30", "ECTO")
comp_stages = do.call(rbind, t)
comp_stages = .grep(comp_stages, seen)

.plot_heatmap = function(PVl){
    keep_mirna = apply(th[th$Var2==PVl,], 1, function(x){
    name = paste0(PVl,th[1])
    k = unique( (comp_stages %>% filter(pval==x[2] & logFC==x[1]))[,"gene"])
    aheatmap(rlog_Mat[k,], labRow = NA, labCol = NA, 
             main = paste0("FDR:",x[2], " logFC:",x[1]), Rowv=FALSE, Colv=FALSE,
             distfun =  "spearman", hclustfun = "ward", legend = FALSE,
             annCol = metadata[colnames(rlog_Mat),"Diffname_short",drop=F])
  })

}

par(mfrow=c(2,2))
.plot_heatmap(0.01)  
.plot_heatmap(0.05)  
.plot_heatmap(0.1)  


```

```{r store, echo=FALSE, eval=FALSE}
parentId = "syn4594275"
activityName = 'Different cutoffs for the DE'
ALL_USED_IDs <- c(mirna_pval_obj, mirna_logfc_obj)

library(rGithubClient)
# Github link
thisFileName = "Different_cutoffs.Rmd"
thisRepo <- getRepo(repository = "lpantano/pcbc_c4_analysis", 
                    ref="branch", 
                    refName='master')

thisFile <- getPermlink(repository = thisRepo,
                        repositoryPath=paste0('code/Rmd/', thisFileName))

CODE <- File(thisFileName,name = activityName,parentId = parentId)
CODE <- synStore(CODE, used = ALL_USED_IDs,activityName=activityName, executed=thisFile)

```

```{r render-synapsy, eval=FALSE, echo=FALSE}
devtools::source_gist("6117476")
knit2synapse("~/repos/pcbc_c4_analysis/code/Rmd/Different_cutoffs.Rmd",
             owner=CODE@properties$id,
             overwrite=TRUE)
```
