---
title: "miRNA-mRNA network analysis with mixed effect modilng"
author: "Lorena Pantano"
date: "`r date()`"
output: html_document
---

```{r knitr,eval=F,echo=F}
library(rmarkdown)
library(knitrBootstrap)
render("~/repos/pcbc_c4_analysis/code/Rmd/miRNA_mRNA_network.Rmd", output_dir = "~/Dropbox/Public/hsph/pcbc")
```


Find miRNA-mRNA networks using limma:voom to detect DE features and miRBD v5 for miRNA-mRNA pairs.


```{r load}
library(tidyr)
library(dplyr)
library(knitr)
library(synapseClient)
library(limma)
library(reshape2)
library(gplots)
library(ggplot2)
library(RColorBrewer)
library(pheatmap)

options(bitmapType = 'cairo')

synapseLogin()

knitr::opts_chunk$set(tidy=TRUE, highlight=TRUE, dev="png", fig.width=9,fig.heigh=9,
               cache=FALSE, highlight=TRUE, autodep=TRUE, warning=FALSE, error=FALSE,
               message=FALSE, prompt=TRUE, comment='', fig.cap='', bootstrap.show.code=FALSE, echo=FALSE)

mirna_pval_obj =  synGet(id="syn4596141")
mirna_pval = read.table(mirna_pval_obj@filePath, check.names = F, header=T, row.names=1, sep="\t")
mirna_logfc_obj =  synGet(id="syn4595547")
mirna_logfc = read.table(mirna_logfc_obj@filePath, check.names = F, header=T, row.names=1, sep="\t")

mrna_logfc_obj =  synGet(id="syn4484226")
mrna_logfc = read.table(mrna_logfc_obj@filePath, check.names = F, header=T, row.names=1, sep="\t")
mrna_pval_obj =  synGet(id="syn4484228")
mrna_pval = read.table(mrna_pval_obj@filePath, check.names = F, header=T, row.names=1, sep="\t")


metadata_id = "syn3219876"
metadata_obj = synTableQuery(paste('SELECT * FROM',metadata_id,sep=' '))
metadata = metadata_obj@values %>% distinct(UID)
row.names(metadata) = metadata$UID
stages =  gsub("-", "", metadata$Diffname_short)
metadata$Diffname_short = as.character(stages)

mirna_rlog_obj =  synGet(id="syn4595977")
rlog_Mat = read.table(mirna_rlog_obj@filePath, check.names = F, header=T, row.names=1, sep="\t")

mrna_metadata_id = "syn3156503"
mrna_metadata_obj = synTableQuery(paste('SELECT * FROM',mrna_metadata_id,sep=' '))
mrna_metadata = mrna_metadata_obj@values %>% distinct(UID)
row.names(mrna_metadata) = mrna_metadata$UID
stages =  gsub("-", "", mrna_metadata$Diffname_short)
mrna_metadata$Diffname_short = as.character(stages)

mrna_rlog_obj =  synGet(id="syn4483934")
rlog_mrna = read.table(mrna_rlog_obj@filePath, check.names = F, header=T, row.names=1, sep="\t")


# meth_metadata_id = "syn3156828"
# meth_metadata_obj = synTableQuery(paste('SELECT * FROM',meth_metadata_id,sep=' '))
# meth_metadata = meth_metadata_obj@values %>% distinct(UID)
# row.names(meth_metadata) = meth_metadata$UID
# stages =  gsub("-", "", meth_metadata$Diffname_short)
# meth_metadata$Diffname_short = as.character(stages)
# 
# meth_rlog_obj = synGet(id="syn4487642")
# rlog_meth = read.table(meth_rlog_obj@filePath, check.names = F, header=T, row.names=1, sep="\t")


meth_mirna_map_id = "syn4895962"
meth_mirna_map_obj = synGet(meth_mirna_map_id)
meth_mirna_map = read.table(meth_mirna_map_obj@filePath, header=F, sep="\t")
names(meth_mirna_map) = c("meth", "mirna")

meth_pval_id = "syn4596513"
meth_pval_obj = synGet(meth_pval_id)
meth_pval = read.table(meth_pval_obj@filePath, header=TRUE, sep ="\t", row.names = 1)

meth_fc_id = "syn4596511"
meth_fc_obj = synGet(meth_fc_id)
meth_fc = read.table(meth_fc_obj@filePath, header=TRUE, sep ="\t", row.names = 1)

.grep <- function(df, comp){
  idx = rowSums( sapply(paste0('_',comp,'_'), function(x)grepl(x, names(df))) ) > 0
  df[,idx]
}

.convert_to_ensembl <- function(m){
  require(biomaRt)
  ensembl = useMart('ensembl', dataset = "hsapiens_gene_ensembl")
  df = getBM(attributes=c("ensembl_gene_id", "external_gene_name"), filters="external_gene_name", values=as.character(rownames(m)), mart=ensembl)
  df = df %>% filter(ensembl_gene_id != "")
  new_m = m[as.character(df$external_gene_name),]
  row.names(new_m) = as.character(df$ensembl_gene_id)
  return(new_m)
}

.convert_to_ensembl_vector <- function(m){
  require(biomaRt)
  ensembl = useMart('ensembl', dataset = "hsapiens_gene_ensembl")
  df = getBM(attributes=c("ensembl_gene_id", "external_gene_name"), filters="external_gene_name", values=as.character(m), mart=ensembl)
  idx = match(as.character(m), df$external_gene_name)
  names = df$ensembl_gene_id[idx]
  names
}


format_table = function(m_logfc, m_pval, FC=2, PVAL=0.01, map=FALSE){
    if (map){
      m_logfc = .convert_to_ensembl(m_logfc)
      m_pval = .convert_to_ensembl(m_pval)
    }
    m_logfc = .grep(m_logfc, comp)
    m_pval = .grep(m_pval, comp)
    
    m_sig = lapply(names(m_pval), function(x){
      
      sig = m_pval[,x] < PVAL & abs(m_logfc[,x]) > FC
      sig[is.na(sig)] = FALSE
      if (sum(sig)>0){
        dt = data.frame(fc=m_logfc[sig, x],gene=rownames(m_logfc)[sig])
      
        dt$cont = x
        return(dt)
      }
    })
    do.call(rbind, m_sig)
}


comp = c("DE", "SC", "MESO5", "ECTO", "EB", "MESO15", "MESO30")

mirna_tab = format_table(mirna_logfc, mirna_pval, FC=2)
meth_tab = format_table(meth_fc, meth_pval, FC=0.2, PVAL=0.05)
mrna_tab = format_table(mrna_logfc, mrna_pval, FC=2, map=TRUE)

meth_gene_obj = synGet(id="syn2775237")
meth_gene = read.table(meth_gene_obj@filePath, check.names = F, header=T, sep="\t")

target_obj = synGet(id="syn3461627")
target = read.table(target_obj@filePath, check.names = F, header=F, sep="\t")
names(target) = c("mirna","gene")
target = target[!duplicated(target),]


meth_gene_obj = synGet(id="syn2775237")
meth_gene = read.table(meth_gene_obj@filePath, check.names = F, header=T, sep=",")
meth_gene = meth_gene[!duplicated(meth_gene),]
meth_gene$ensembl = .convert_to_ensembl_vector(meth_gene$UCSC_RefGene_Name)
```

# Detect networks

For mRNA and miRNA using adjusted pval < 0.01 and abs(logFC) > 2 and for methylation, abs(logFC) > 0.2 and FDR < 0.05 to get more interactions.

```{r create-network}
# MiRNA vs Gene
network_list = list()
for (contrast in intersect(unique(mrna_tab$cont), unique(mirna_tab$cont)) ){
    genes_de = mrna_tab %>% filter(cont == contrast )
    mirna_de = mirna_tab %>% filter(cont == contrast)
    target_de = target %>% filter(gene %in% genes_de$gene)  %>% filter(mirna %in% mirna_de$gene)
    idx_mirna = match(target_de$mirna, mirna_de$gene)
    idx_mrna = match(target_de$gene, genes_de$gene)
    target_de[!is.na(idx_mirna),"mirna_fc"] = mirna_de$fc[idx_mirna[!is.na(idx_mirna)]]
    target_de[!is.na(idx_mrna),"mrna_fc"] = genes_de$fc[idx_mrna[!is.na(idx_mrna)]]
    network = target_de %>% filter(!is.na(mirna_fc) & !is.na(mrna_fc) & mirna_fc * mrna_fc < 0)
    if (dim(network)[1]>0){
      network_list[[contrast]] = network %>% mutate(contrast = contrast)
    }
}

# Methylation vs miRNA
network_meth_list = list()
for (contrast in intersect(unique(meth_tab$cont), unique(mirna_tab$cont)) ){
    genes_de = meth_tab %>% filter(cont == contrast )
    mirna_de = mirna_tab %>% filter(cont == contrast)
    target_de = meth_mirna_map %>% filter(meth %in% genes_de$gene & mirna %in% mirna_de$gene)
    idx_mirna = match(target_de$mirna, mirna_de$gene)
    idx_mrna = match(target_de$meth, genes_de$gene)
    target_de[!is.na(idx_mirna),"mirna_fc"] = mirna_de$fc[idx_mirna[!is.na(idx_mirna)]]
    target_de[!is.na(idx_mrna),"meth_fc"] = genes_de$fc[idx_mrna[!is.na(idx_mrna)]]
    network = target_de %>% filter(!is.na(mirna_fc) & !is.na(meth_fc) & mirna_fc * meth_fc < 0)
    if (dim(network)[1]>0){
      network_meth_list[[contrast]] = network %>% mutate(contrast = contrast)
    }
}

# Methylation vs genes
network_meth_gene_list = list()
for (contrast in intersect(unique(meth_tab$cont), unique(mrna_tab$cont)) ){
    meth_de = meth_tab %>% filter(cont == contrast )
    gene_de = mrna_tab %>% filter(cont == contrast)
    target_de = meth_gene %>% filter(Name %in% meth_de$gene & ensembl %in% gene_de$gene) %>% dplyr::select(Name, ensembl)
    idx_gene = match(target_de$ensembl, gene_de$gene)
    idx_meth = match(target_de$Name, meth_de$gene)
    target_de[!is.na(idx_gene),"gene_fc"] = gene_de$fc[idx_gene[!is.na(idx_gene)]]
    target_de[!is.na(idx_meth),"meth_fc"] = meth_de$fc[idx_meth[!is.na(idx_meth)]]
    network = target_de %>% filter(!is.na(gene_fc) & !is.na(meth_fc) & gene_fc * meth_fc < 0)
    if (dim(network)[1]>0){
      network_meth_gene_list[[contrast]] = network %>% mutate(contrast = contrast)
    }
}

```

# Numbers of interaction in all comparisons

## For mRNA-miRNA

```{r summary, results='asis'}
kable(as.data.frame(sapply(network_list, nrow)))
```

## For mRNA-meth

```{r summary-mrna-meth, results='asis'}
kable(as.data.frame(sapply(network_meth_gene_list, nrow)))
```

## For methylation-miRNA

```{r summary-meth, results='asis'}
kable(as.data.frame(sapply(network_meth_list, nrow)))
```

## Common interactions for mRNA-miRNA

```{r summary-unique, results='asis'}
dd = do.call(rbind, network_list) 
dd$id = paste0(dd$mirna,"->",dd$gene, "@", dd$contrast)
dd$interaction = paste0(dd$mirna,"->",dd$gene)
dd = dd %>% filter(!duplicated(c(id)))
dd_times = dd %>% group_by(interaction) %>% dplyr::summarise(times = n()) %>% dplyr::arrange(desc(times))
dd_complete = merge(dd, dd_times, by="interaction") %>% dplyr::arrange(desc(times))

kable(head(merge(dd, dd_times, by="interaction") %>% filter(times>1) %>% dplyr::arrange(desc(times))))

```

## methylation - miRNA - mRNA interaction

Number of genes with methylation-miRNA-mRNA interactions

```{r c3-network}
meth_mirna_res = do.call(rbind,network_meth_list) %>% mutate(id=paste0(mirna,"@", contrast))
mrna_mirna_res = do.call(rbind,network_list) %>% mutate(id=paste0(mirna, "@",contrast))
meth_mrna_res = do.call(rbind,network_meth_gene_list) %>% mutate(id=paste0(Name,"@", contrast))

c3_res = merge(mrna_mirna_res, meth_mirna_res, by ='id')

kable(as.data.frame(table(c3_res$contrast.x)))

```

# Importants miRNAs for each state

The idea is to take the most important miRNAs that repressed a set of genes in each state
that we assume are the ones controling tahta the state remains like it is.

So we are lookgin really at UP miRNAs that control the down-regulated genes that
are unique to this state. So, we asume that changes in these genes would start
a differentiation to other state. As a note, downregulated doesn't mean low expressed,
some of them are low expressed, others are lower than other state.

Steps for each state:

* take only unique downregulated genes for this state. That means that those
genes are differentially down-regulated from any other state to this one
* counts the miRNAs that interact those genes and the ones that are UP are
the candidates that control those genes that are down. 
> * take the miRNAs that target more genes (top 10, but we can add some stats here to chose better)
* take the miRNAs that target more genes. miRNAs were sorted by total number of targeted genes. Sliding windows of 5 miRNAs runs through the list and the ratio between the number of targeted genes from the previous windows was calculated. All miRNAs above the last ratio > 20% are taking as the most important. As well, only of those that targeted < 5% of the total were discarted to avoid noise due to low number of genes. 
* group these miRNAs if they have many common targets (set to > 50% in common)
* remove miRNAs that targets < 10% of all the genes and are not group to any other miRNA
* remove genes that after the 2 previous step have no miRNAs targeting them
* plot mRNA expression (scale by row) in all states and annotate at the left 
which miRNAs are regulating them (black:yes, white:no)

```{r co-profile}

states = c("DE", "SC", "EB", "ECTO", "MESO5", "MESO15", "MESO30")

.name2entrez = function(x, out="list"){
    require(org.Hs.eg.db)
    symbol = AnnotationDbi::select(org.Hs.eg.db, as.character(x), "ENTREZID", keytype="ENSEMBL")
    symbol = symbol %>% distinct(ENTREZID)
    if (out=="table")
        return(symbol)
    symbol$ENTREZID[!is.na(symbol$ENTREZID)]
}


mrna_mirna_res$symbol = .name2symbol(mrna_mirna_res$gene, out='table')
meth_mrna_res$symbol = .name2symbol(meth_mrna_res$ensembl, out='table')

.list = lapply(states, function(state){
  
  idx = grepl(state, unique(mrna_mirna_res$contrast))
  .genes=lapply(unique(mrna_mirna_res$contrast)[idx], function(comp){
    .split = unlist(strsplit2(comp,"_"))
    .side = which(.split==state)
    if (.side==3){
      .res_mirna = mrna_mirna_res %>% filter(contrast==comp & mrna_fc < 0)
      .res_meth = meth_mrna_res %>% filter(contrast==comp & gene_fc < 0 )
    }else{
      .res_mirna = mrna_mirna_res %>% filter(contrast==comp & mrna_fc > 0)
      .res_meth = meth_mrna_res %>% filter(contrast==comp & gene_fc > 0 )
    }
    return(list(mirna=unique(as.character(.res_mirna$gene)), meth=as.character(.res_meth$ensembl)))
  })
  
  
  .genes_mirna_counts = data.frame(genes=unlist(sapply(.genes, function(x){x$mirna}))) %>% group_by(genes) %>% summarise(total=n()) %>% filter(total>0)
  .genes_meth_counts = data.frame(genes=unlist(sapply(.genes, function(x){x$meth}))) %>% group_by(genes) %>% summarise(total=n()) %>% filter(total>0) # remove 3 to 0 to create 1 version
  
  # print(.genes_mirna_counts)
  .genes_mirna = as.character(.genes_mirna_counts$genes)
  .genes_meth = as.character(.genes_meth_counts$genes)
  .all = unique(c(.genes_mirna, .genes_meth))
  .mirna = mrna_mirna_res %>% filter(grepl(state,contrast) & gene %in% .all) %>% 
    distinct(gene,mirna) %>% group_by(mirna) %>% summarise(times=n()) %>% arrange(desc(times))
  .meth = meth_mrna_res %>% filter(grepl(state,contrast) & ensembl %in% .all) %>% 
    distinct(ensembl,Name) 
  
  .entrez = .name2entrez(.all)
  return(list(entrez=.entrez, ensembl=.all, mirna=.mirna, meth=.meth, genes_mirna=.genes_mirna, genes_mirna_counts=.genes_mirna_counts, genes_meth_counts=.genes_meth_counts))  

})

names(.list) = states


.name2symbol = function(x, out="list"){
    require(org.Hs.eg.db)
    symbol = AnnotationDbi::select(org.Hs.eg.db, as.character(x), "SYMBOL", keytype="ENSEMBL")
    symbol = symbol %>% distinct(SYMBOL)
    if (out=="table"){
        idx = match(as.character(x), symbol$ENSEMBL)
        symbol = symbol[idx,]
        symbol$SYMBOL[is.na(symbol$ENSEMBL)] = symbol$ENSEMBL[is.na(symbol$ENSEMBL)]
        return(symbol$SYMBOL)
      }
    symbol$SYMBOL[!is.na(symbol$SYMBOL)]
}

# .up_genes = as.character(unlist(unique(lapply(.list, function(x){x$ensembl}))))
```

```{r reduce-fn}
# this_state = 'MESO15'
.top_mirna = unique(as.character(unlist(lapply(states, function(state){
    unlist(.list[[state]]$mirna[1:5,"mirna"])
  }))))


.get_mirna_mrna_ma = function(.toy, min_total=0.1){
  .seen = vector()
  .sort_mirna = sort(colSums(.toy>0), decreasing = T)
  .pairs = vector("list", length=length(.sort_mirna))
  for (nc1 in 1:length(.sort_mirna)){
    for (nc2 in 1:length(.sort_mirna)){
      if (nc1 != nc2 & sum(.seen==nc2)==0){
        .name1 = names(.sort_mirna)[nc1]
        .name2 = names(.sort_mirna)[nc2]
        .these1 = rownames(.toy)[.toy[,.name1]>0]
        .these2 = rownames(.toy)[.toy[,.name2]>0]
        .min = min(length(.these1), length(.these2))
        .common = length(intersect(.these1, .these2))
        if ( (.common / .min) > 0.50 ){
          .seen = unique(c(.seen , nc2, nc1))
          .pairs[[nc1]] = c(.pairs[[nc1]], nc2)
        }
      }
    }
  }
  
  .others = c()
  for (nc1 in 1:length(.sort_mirna)){
    if (sum(.seen==nc1)==0 ){
      .these1 = rownames(.toy)[.toy[,nc1]>0]
      if ( length(.these1) / nrow(.toy) >= min_total ){
        .pairs[[nc1]] = c(.pairs[[nc1]], nc1)
      }#else{
      #.others = c(.others, nc1)
      #}
      .seen = c(.seen , nc1)
    }
  }
  .toyb <- .toy
  .toyb_red = lapply(1:length(.sort_mirna), function(x){
    if ( ! is.null(.pairs[[x]]) ){
      .join = unique(c(names(.sort_mirna)[x],names(.sort_mirna)[.pairs[[x]]]))
      .name = gsub('hsa-', '', gsub("miR-", "", paste0(.join, collapse = ",")))
      .news = rowSums(.toyb[,.join, drop=F])
      # .news[.news>0] = 1
      if ( sum(.news) >= nrow(.toyb) * min_total)
        return(list(v=.news, n=.name, real=.join))
    }
  })

  .used = unique(unlist(lapply(.toyb_red, function(x){
    x$real
  })))
  .names = unlist(lapply(.toyb_red, function(x){
    x$n
  }))
  .toyb_red = do.call(cbind, lapply(.toyb_red, function(x){
    x$v
  }))
  colnames(.toyb_red) = .names
  .toyb_red = as.data.frame(.toyb_red)

  .toy_hc = rownames(.toyb_red)
  if (ncol(.toyb_red) > 1){
    cat("columns", ncol(.toyb_red))
    .toy_hc = hclust(dist(t(.toyb_red), method='binary'), method="ward.D2")
  }
  

  .toy_red = .toyb_red
  .toy_red[.toyb_red==0] = "None"
  .toy_red[.toyb_red!=0] = "Target"
  
  .ann_colors = lapply(colnames(.toy_red), function(x) {c(None="azure", Target="grey")})
  names(.ann_colors) = colnames(.toy_red)
  return(list(bin=.toyb_red, cat=.toy_red, cols=.ann_colors, hc=.toy_hc, used =.used))
}
```

```{r reduce}
all_tab_mrna = data.frame()
all_tab_ann = data.frame()
.states_lvs = factor(states, levels = c("SC", "DE", "MESO5",  "ECTO", "EB","MESO15", "MESO30"))

all_info = lapply(as.character(levels(.states_lvs)), function(this_state){
  print(this_state)
  # Select genes down only here with some interaction with mirna or methylation
  .down_genes = as.character(unique(unlist(.list[[this_state]]$ensembl)))
  .all_genes_mi = (mrna_mirna_res %>% filter(!grepl(this_state, contrast)))[,"gene"]
  .all_genes_meth = (meth_mrna_res %>% filter(!grepl(this_state, contrast)))[,"ensembl"]
  .down_uni = setdiff(setdiff(.down_genes, .all_genes_mi), .all_genes_meth)
  
  # Select top miRNAs that target most genes
  .top_mirna = mrna_mirna_res %>% filter(gene %in% .down_uni & grepl(this_state, contrast)) %>% filter(!is.na(symbol)) %>% group_by(mirna) %>% summarise(times=n()) %>% arrange(desc(times))

  .top_mirna = .top_mirna %>% filter(times>length(.down_uni)*0.05)
  .n = nrow(.top_mirna)
  if (.n > 5){
  .changes = (.top_mirna$times[1:(.n-5)] / .top_mirna$times[3:(.n-3)])
  .cutoff = sort(which(.changes > 1.20), decreasing = T)[1]
  }else{.cutoff=.n}
  # .top_mirna$mirna[1:.cutoff]
  # %>% filter(times > length(.down_uni)*0.05)
  .top_mirna = as.character(unique(unlist(.top_mirna$mirna[1:.cutoff])))
  
  # Filter intaractions only with selected genes and in this state
  .filtered_mi = mrna_mirna_res %>% filter(mirna %in% .top_mirna & gene %in% .down_uni & grepl(this_state, contrast)) %>% filter(!is.na(symbol))
  .filtered = .name2symbol(.down_uni)
  
  # Expression of selected genes and translate to symbol
  .mrna_exp = rlog_mrna[as.character(unique(.filtered)),]
  .mirna_exp = rlog_Mat[as.character(unique(.filtered_mi$mirna)),]
  .mirna_genes = as.character(unique(.filtered_mi$symbol))
  
  # Calculate median by group
  .median_by_group = function(mat, meta){
    meta = meta[colnames(mat),]
    .merge = do.call(cbind, lapply(states, function(state){
      .these = as.character((meta  %>% filter(Diffname_short == state))[,"UID"])
      
      .summarize = apply(mat[,.these],1,median)
    }))
    colnames(.merge) = states
    .merge[!is.na(rowSums(.merge)),]
    
  }
  mirna_keep = .median_by_group(.mirna_exp, metadata)
  mrna_keep = .median_by_group(.mrna_exp, mrna_metadata)
  
  
  .hc = function(mat){
    hclust(as.dist(1-cor(t(mat))), method="ward.D2")
  }
  .plot_heatmap = function(mat, lvls){
    ._plot = melt(as.data.frame(mat) %>%  mutate(feature=rownames(mat)))
    ._plot$mirna = factor(._plot$feature, levels=levels(lvls))
    ._plot$variable = factor(._plot$variable, levels=levels(.states_lvs))
    
    p = ggplot(._plot, aes(x=variable, y=feature, fill=value)) +
      geom_tile() +
      scale_fill_gradient2(low="blue", mid = "white", high = "orange", midpoint = 3)
    print(p)
  }
  
  
  # Prepare matrix with genes/mirna interaction
  # Final matrix with 0 if mirna doesn't target gene 
  # or >0 if it does in the different states
  .ma_genes_mirna = matrix(ncol=nrow(.mirna_exp), nrow=length(.mirna_genes))
  rownames(.ma_genes_mirna)= .mirna_genes
  colnames(.ma_genes_mirna)= rownames(.mirna_exp)
  .ma_genes_mirna[is.na(.ma_genes_mirna)] = 0
  .void = apply(.filtered_mi, 1, function(x){
    x = as.character(x)
    if ( sum(x[7]==rownames(.ma_genes_mirna))>0 & sum(x[1]==colnames(.ma_genes_mirna))>0 )
      .ma_genes_mirna[x[7], x[1]] <<- .ma_genes_mirna[x[7], x[1]] + 1
  })
  
  # print(colSums(.toy==1)/nrow(.toy))
  # Merge miRNAs into families and common targets for viz porpuses
  .mirna_mrna_ma = .get_mirna_mrna_ma(.ma_genes_mirna,min_total = 0.1)  
  .ann_bin_red = .mirna_mrna_ma$bin # binary matrix
  .ann_red = .mirna_mrna_ma$cat # character category matrix
  .ann_colors = .mirna_mrna_ma$cols # colores for the plot
  .ann_hc = .mirna_mrna_ma$hc # orger of mirnas groups
  
  # tt = .list[[this_state]]$genes_mirna_counts
  # tt$symbol = .name2symbol(tt$genes, out='table')
  # .toyb_red_norm = sweep(.toyb_red, 1, unlist(tt[match(rownames(.toyb_red), tt$symbol),"total"]), "/")
  # Define annotation for miRNAs for each gene
  .ann_df=as.data.frame(.ann_red[,.ann_hc$order])

  # Get genes finally regulated by miRNAs
  .regulated = intersect(rownames(.ann_bin_red)[rowSums(.ann_bin_red)>0], rownames(mrna_keep))
  
  # Select genes only regulated by miRNAs to plot first
  .regulated_meth = intersect(.list[[this_state]]$meth$symbol, rownames(mrna_keep))
  .regulated = setdiff(.regulated, .regulated_meth)
  .ma = mrna_keep[.regulated, ]
  .ma_hc = hclust(as.dist(1-cor(t(.ma))^2))
  .all_ann = rbind(.ann_df[.regulated,], .ann_df[.regulated_meth,])
  
  # Select genes reguated by miRNA/methylation to plot second
  if (length(.regulated_meth) > 0){
    .ma_meth = mrna_keep[.regulated_meth, ]
    .ma_hc_meth = hclust(as.dist(1-cor(t(.ma_meth))^2))
    .all_ma = rbind(.ma[.ma_hc$order ,levels(.states_lvs)], .ma_meth[.ma_hc_meth$order ,levels(.states_lvs)])
  }else{
    .all_ma = .ma[.ma_hc$order ,.states_lvs]
  }
  print(dim(.all_ma))
  # Calculate genes not regulated by any
  .unkown = nrow(mrna_keep) - nrow(.all_ma)
  
  # Define colors for methylated genes
  .all_ann[.regulated_meth, "methylated"] = "Target"
  .ann_colors[['methylated']] = c(None="white", Target="grey")
  
  pheatmap(.all_ma, show_rownames = FALSE, cluster_cols = FALSE, cluster_rows = FALSE, annotation_row = .all_ann, annotation_colors = .ann_colors, scale = 'row', annotation_legend = FALSE)

  list(ma=.all_ma, ann=.all_ann, col=.ann_colors, 
       used=.mirna_mrna_ma$used, 
       meth_used = .regulated_meth, 
       meth_probes = .list[[this_state]]$meth %>% filter(symbol %in% .regulated_meth),
       not_explained = .unkown)
})

names(all_info) = levels(.states_lvs)
```


## All together

basically took all previous heatmaps and paste one after another. 
the miRNA annotation is redone here. Just take all miRNAs in the previous heatmaps
and group again together according to targets, but without removing anyone. Here I just added
some miRNA-gene annotation at the left since we have more genes here.

```{r plot}
# Get all data to create all state together
all_ma = do.call(rbind, lapply(all_info, function(x){x$ma}))
all_used = unique(unlist(lapply(all_info, function(x){x$used})))
.filtered_mi = mrna_mirna_res %>% filter(mirna %in% all_used & symbol %in% as.character(rownames(all_ma))) %>% filter(!is.na(symbol))

# Prepare matrix with genes/mirna interaction
# Final matrix with 0 if mirna doesn't target gene 
# or >0 if it does in the different states
.ma_genes_mirna = matrix(ncol=length(all_used), nrow=nrow(all_ma))
rownames(.ma_genes_mirna)= rownames(all_ma)
colnames(.ma_genes_mirna)= all_used
.ma_genes_mirna[is.na(.ma_genes_mirna)] = 0
.void = apply(.filtered_mi, 1, function(x){
  x = as.character(x)
  if ( sum(x[7]==rownames(.ma_genes_mirna))>0 & sum(x[1]==colnames(.ma_genes_mirna))>0 )
    .ma_genes_mirna[x[7], x[1]] <<- .ma_genes_mirna[x[7], x[1]] + 1
})

# Merge miRNAs into families and common targets for viz porpuses
.mirna_mrna_ma = .get_mirna_mrna_ma(.ma_genes_mirna, min_total = 0)  
.ann_bin_red = .mirna_mrna_ma$bin
.ann_red = .mirna_mrna_ma$cat
ann_colors = .mirna_mrna_ma$cols
.ann_hc = .mirna_mrna_ma$hc

ann_df = as.data.frame(.ann_red)

# Color dark if miRNA is top in the list of that state or gray if not
.void = lapply(all_info, function(x){
  .m = x$used
  .m = gsub("miR-","",gsub("hsa-","",.m))
  .all = unlist(sapply(names(ann_df), function(x){ unlist(strsplit2(x,",")) }))
  .all_names = unlist(sapply(names(ann_df), function(x){ rep(x,length(unlist(strsplit2(x,",")))) }))
  
  idx = match(.m, .all)
  .seen = .all_names[idx]
  .seen = unique(sapply(.seen,function(x){
    if (grepl("-",x))
      x = gsub("[0-9]$","",x)
    x
    }))

  for (.m in unique(.seen)){
    .ma = ann_df[rownames(x$ma),.m] == "Target"
    if (sum(.ma)>0)
      ann_df[rownames(x$ma),.m][.ma] <<- "Top10"
  }
    
})

# Clustering annotation only considering Top miRNAs in each state
.ann_bin_red[as.matrix(ann_df)=="Target"] = 0
.hc = hclust(dist(t(.ann_bin_red), method='binary'), method="ward.D2")
ann_df = ann_df[, .hc$order]

# Add methylation information
.all_meth = unlist(sapply(all_info, function(x){x$meth_used}))
ann_df$methylated = "None"
ann_df[.all_meth, "methylated"] = "Target"
ann_colors[['methylated']] = c(None="white", Target="black")

# Choose blak as top miRNA or methylated
ann_colors = lapply(ann_colors,function(x){  c(x, Top10="black") })

# Merge miRNAs in a same name (i.e a/b/c/d ... into 1)
short_minra_names=sapply(names(ann_df), function(x){
  .items = as.vector(unlist(strsplit2(split = ",",x)))
  .seen = c()
  .new = list()
  for (i in .items){
    .let = ""
    if (grepl("let", i)){ .let="let-"}
    i = gsub("let-", "", i)
    if (i=="methylated"){return(i)}
    .mir = as.vector(unlist(strsplit2(i,split = "-")))[1]
    .side = as.vector(unlist(strsplit2(i,split = "-")))[2]
    .names = as.vector(unlist(strsplit2(.mir,split = "[a-z]")))
    .fam = as.vector(unlist(strsplit2(.mir,split = "[0-9]")))
    for (n in .names[1]){
      n_w_side = gsub("-NA","",paste0(.let, n,"-",.side))
      if (n != "" & (n_w_side %in% .seen)){
        .new[[n_w_side]] = unique(c(.new[[n_w_side]], .fam))
      }else if (n != "" & !(n_w_side %in% .seen)){
        .new[[n_w_side]] = .fam
        .seen = c(.seen, n_w_side)
      }
    }
    
  }
  .fn = function(x){x!=""}
  .new = sapply(names(.new), function(x){
    .new[[x]]=Filter(.fn,.new[[x]])
    if (length(.new[[x]])>0){
      return(paste0(x,"[",paste0(.new[[x]], collapse = "|"),"]"))
    }else{
      return(x)
    }
  })
  paste0(unlist(.new), collapse = ",")
})

# Mark spaces between states
gaps = cumsum(sapply(all_info, function(x){nrow(x$ma)}))
cols = colorRampPalette(rev(c("darkorange3","lightyellow","deepskyblue4")))(100)
pheatmap(all_ma, color = cols, show_rownames = FALSE, cluster_cols = FALSE, cluster_rows = FALSE, scale='row', annotation_legend = FALSE, annotation_row = ann_df, annotation_colors = ann_colors, gaps_row=gaps)

# Genes not explained
sapply(all_info,function(x){
  x$not_explained/(nrow(x$ma)+x$not_explained)
})

```


```{r mirna-exp-heatmap}
# meth_exp = sapply(as.character(levels(.states_lvs)), function(this_state){
#       print(this_state)
#       if (this_state %in% c("MESO15", "MESO30") ){
#         return(NA)
#       }
#       cg_probes = as.character(unique(all_info[[this_state]]$meth_probes[,"Name"]))
#       median(rowMedians(as.matrix(rlog_meth[cg_probes, meth_metadata[meth_metadata$UID %in% names(rlog_meth) & meth_metadata$Diffname_short==this_state, "UID"], drop=FALSE])))
# })

# Get medium expression by miRNA and state
mirna_exp = lapply(names(ann_df), function(x){
  .m = gsub("miR-","",gsub("hsa-","",rownames(rlog_Mat)))
  .all = unlist(sapply(x, function(x){unlist(strsplit2(x,","))}))

  idx = match(.all, .m)
  if (x != "methylated"){
    .exp = sapply(as.character(levels(.states_lvs)), function(this_state){
      median(rowMedians(as.matrix(rlog_Mat[idx, metadata[metadata$UID %in% names(rlog_Mat) & metadata$Diffname_short==this_state, "UID"], drop=FALSE])))
     })
    return(.exp)
  }
})

mirna_exp = matrix(unlist(mirna_exp), ncol=7, byrow = TRUE)
colnames(mirna_exp) = levels(.states_lvs)
rownames(mirna_exp) = as.vector(unlist(short_minra_names))[1:nrow(mirna_exp)]

```

```{r gplot-final}
library(grid)
# miRNA heatmap
.ma = apply(mirna_exp,1,scale)
rownames(.ma) = colnames(mirna_exp)
.ma = as.data.frame(.ma[rev(levels(.states_lvs)), rev(rownames(mirna_exp))])

.ma_melt =  melt(.ma %>% mutate(methylated=NA, state=rownames(.ma)), id.vars = "state")
.ma_melt$state = factor(.ma_melt$state, levels(.states_lvs))
.ma_melt$variable = factor(.ma_melt$variable, c("methylated",colnames(.ma)))

gmirna = ggplot(.ma_melt, aes(x=variable, y=state, fill=value)) +
  geom_tile() + 
  geom_vline(xintercept=seq(0.5,(ncol(.ma)+0.5),1), color="white", size=0.5) + 
  scale_fill_gradient2(na.value = "white",guide=FALSE, high="darkorange3", low="deepskyblue4", mid = "lightyellow", midpoint = 0) +
  theme_minimal() + theme(plot.margin = unit(c(-1,-0.5,0,0), "lines"),
                          panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
                          axis.ticks = element_line(color="white"), axis.text.y=element_blank(),             
                          axis.text.x = element_text(angle=90, size=9, hjust = 1, vjust=0.5)) +
  xlab("") + ylab("")


# mRNA heatmap
.ma_mrna = t(apply(all_ma,1,scale))
colnames(.ma_mrna) = colnames(all_ma)
.mrna_melt = melt(as.data.frame(.ma_mrna) %>% mutate(gene=nrow(all_ma):1),id.vars = c("gene"))
.mrna_melt$variable = factor(.mrna_melt$variable, levels(.states_lvs))

gmrna = ggplot(.mrna_melt, aes(y=gene, x=variable, fill=value)) +
  geom_tile() + 
  scale_fill_gradient2(guide=FALSE, high="darkorange3", low="deepskyblue4", mid = "lightyellow", midpoint = 0) +
  geom_hline(yintercept=nrow(.ma_mrna) - gaps, color="white", size=2) + 
  theme_minimal() + theme(plot.margin = unit(c(0,0.5,-0.5,-0.5), "cm"),
                          panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
                          axis.ticks = element_blank(), axis.text.y = element_blank(),
                          axis.text.x = element_blank()) +
  xlab("") + ylab("")

# Annotation mirna/meth with genes
.ma_ann = ann_df
.ma_ann[.ma_ann$methylated=="Target", "methylated"] = "Top10"
.ann_melt = melt(as.data.frame(.ma_ann) %>% mutate(gene=nrow(all_ma):1),id.vars = c("gene"))
.ann_melt$variable = factor(.ann_melt$variable, rev(colnames(ann_df)))

gann = ggplot(.ann_melt, aes(y=gene, x=variable, fill=value)) +
  geom_tile(size=1) + 
  geom_vline(xintercept=seq(0.5,(ncol(.ma)+0.5),1), color="white", size=1) +
  scale_fill_manual(values = c("azure", "grey", "black"), guide=FALSE) +
  geom_hline(yintercept=nrow(all_ma) - gaps, color="white", size=2) + 
  theme_minimal() + theme(plot.margin = unit(c(0,-0.5,-0.5,0), "cm"),
                          panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
                          axis.text.x=element_blank(), axis.ticks = element_blank(), 
                          axis.text.y = element_blank()) +
  xlab("") + ylab("")

# Labels for heatmaps with each state
# select the longest label of the mirna plot to align this one at the bottom
.longest_name = short_minra_names[which.max(as.vector(sapply(short_minra_names,nchar)))[1]]
axis_df = data.frame(x=1:ncol(.ma_mrna), y =1:ncol(.ma_mrna), text=colnames(.ma_mrna))
axisPlot <- ggplot(axis_df, aes(x=x,y=y, fill="white")) +
  geom_tile()+
  geom_text(aes(label = text), size=3) +
  scale_fill_manual(guide=FALSE, values = "azure") +
  xlab(.longest_name) +
  theme(
    plot.margin = unit(c(-1,-0.2,0,-0.9), "cm"),
    plot.background = element_blank(), 
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(), 
    panel.border = element_blank(),
    panel.background = element_blank(),
    axis.title.x = element_text(color="white",angle=90, size=9),
    axis.title.y = element_blank(),
    axis.text.x = element_blank(), 
    axis.text.y = element_blank(),
    axis.ticks = element_blank(),
    axis.line = element_blank()
     )


# Composition
library("gridExtra")
grid.arrange(gann, gmrna, gmirna, axisPlot, padding=0,
        ncol=2, nrow=2, widths=c(4, 3), heights=c(4, 4), clip=TRUE)

pdf("../../data/omics.pdf", paper = "a4r",pointsize = 9, width = 700)
grid.arrange(gann, gmrna, gmirna, axisPlot, padding=0,
        ncol=2, nrow=2, widths=c(4, 3), heights=c(4, 4), clip=TRUE)
dev.off()
```

```{r store, eval=FALSE}
parentId = "syn4596210"
ActivityName = 'miRNA-mRNA network from mixedEffect DE'
library(rGithubClient)
# Github link
thisFileName = "miRNA_mRNA_network.Rmd"
thisRepo <- getRepo(repository = "lpantano/pcbc_c4_analysis", 
                    ref="branch", 
                    refName='master')

thisFile <- getPermlink(repository = thisRepo,
                        repositoryPath=paste0('code/Rmd/', thisFileName))

ALL_USED_IDs = c(mirna_rlog_obj$properties$id, metadata_obj@schema, mrna_logfc_obj@properties$id, mrna_pval_obj@properties$id, mirna_logfc_obj@properties$id, mirna_pval_obj@properties$id, meth_pval_id, meth_fc_id, meth_mirna_map_id, mrna_rlog_obj@properties$id)

CODE <- File(thisFileName,name = ActivityName,parentId = parentId)
CODE <- synStore(CODE, used = ALL_USED_IDs, activityName=ActivityName, executed=thisFile)

write.table(mrna_mirna_res, "../../data/miRNA-mRNA_mixedEffect_netowrk.tsv", quote=F, sep="\t", row.names=F)
res <- File('../../data/miRNA-mRNA_mixedEffect_netowrk.tsv',name = 'miRNA-mRNA network list',parentId = parentId)
res <- synStore(res, used = ALL_USED_IDs, activityName=ActivityName, executed=CODE)

write.table(meth_mirna_res, "../../data/miRNA-meth_mixedEffect_netowrk.tsv", quote=F, sep="\t", row.names=F)
res <- File('../../data/miRNA-meth_mixedEffect_netowrk.tsv',name = 'miRNA-meth network list',parentId = parentId)
res <- synStore(res, used = ALL_USED_IDs, activityName=ActivityName, executed=CODE)

write.table(all_ma, "../../data/miRNA-mRNA_mixedEffect_median_exp_by_group.tsv", quote=F, sep="\t")
res <- File('../../data/miRNA-mRNA_mixedEffect_median_exp_by_group.tsv',name = 'gene median expression by stage',parentId = parentId)
res <- synStore(res, used = ALL_USED_IDs, activityName=ActivityName, executed=CODE)

write.table(ann_df, "../../data/miRNA-mRNA_mixedEffect_gene_annotation.tsv", quote=F, sep="\t")
res <- File('../../data/miRNA-mRNA_mixedEffect_gene_annotation.tsv',name = 'gene annotation by methylation and miRNAs',parentId = parentId)
res <- synStore(res, used = ALL_USED_IDs, activityName=ActivityName, executed=CODE)


res <- File("../../data/omics.pdf",name = 'gene regulation plot',parentId = parentId)
res <- synStore(res, used = ALL_USED_IDs, activityName=ActivityName, executed=CODE)

devtools::source_gist("6117476")
knit2synapse("miRNA_mRNA_network.Rmd",
             owner=CODE@properties$id,
             overwrite=TRUE)
```

